---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Annabelle White"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(lubridate)
library(agricolae)
library(ggplot2)
library(ggthemes)
library(here)
library(corrplot)

Lake.raw <- read.csv(
  file = here('Data','Raw', 'NTL-LTER_Lake_ChemistryPhysics_Raw.csv'),
  stringsAsFactors = T
)

Lake.raw$sampledate <- mdy(Lake.raw$sampledate) # Convert to date objects

#2

starwars <- theme_base() +  # I still like my silly little Star Wars theme
  theme( # Pulled from John's theme template code
    line = element_line(color = "white"), 
    rect = element_rect(fill = "black"),
    text = element_text(color = "yellow"), 
    
    # Modified inheritance structure of line element
    axis.ticks = element_line(color = "yellow"),
    panel.grid.major = element_line(color = "yellow", 
                                    linetype = "dashed")
    )

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: Mean lake temperature in July does not vary over depth in every lake.
Ha: Mean lake temperature in July varies over depth in every lake.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

Lake.pipe <- 
  Lake.raw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na() # Use this instead of na.omit() for tidyverse!

#5

tempdepth.plot <- ggplot(Lake.pipe, aes(x = depth, 
                          y = temperature_C)) + 
  geom_point(size = 0.5, color = "white") + 
  starwars + # Add my theme
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              color = "blue") +
  ylim(0, 35) + # Limit temperature to 35 degrees C

  # Add labels and titles
  xlab("Lake Depth (m)") +
  ylab("Temperature (degrees C)") + 
  ggtitle("July Temperatures by Depth")

print(tempdepth.plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
The figure suggests that temperature and depth are inversely related. As depth increases, average temperatures decrease. Additionally, there is wider variation in temperature at low depths than at high depths. As depth increases, the data show much less distance between high and low temperatures.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

tempdepth.lm <- lm(data = Lake.pipe, temperature_C ~ depth)
summary(tempdepth.lm) # Shows significant negative effect

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
For each 1-meter increase in depth, temperature decreases by 1.94621 degrees C. The result is highly significant, with a p-value less than 2.2e-16, on 9726 degrees of freedom. The model has an adjusted R-squared of 0.7387, suggesting that this only explains about 74% of the variability in the data. There may be other variables affecting temperature besides depth.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 

9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

tempAIC <- lm(data = Lake.pipe, temperature_C ~ year4 + daynum + depth)
step(tempAIC) # Suggests that removing any of these would hurt the model

#10

temp.lm <- lm(data = Lake.pipe, temperature_C ~ depth + daynum + year4)
summary(temp.lm)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
Depth, year, and day together have an adjusted R-squared of 7.411, meaning they explain approximately 0.23% more of the variability in the data than depth alone, with two fewer degrees of freedom.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# Format ANOVA as aov
Lake.anova1 <- aov(data = Lake.pipe, temperature_C ~ lakename)
summary(Lake.anova1)

# Format ANOVA as lm
Lake.anova2 <- lm(data = Lake.pipe, temperature_C ~ lakename)
summary(Lake.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
Yes. Both ANOVA found significant difference in mean July temperatures between the lakes, with a p-value less than 2e-16.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

tempdepth.plot2 <- ggplot(Lake.pipe, aes(x = depth, 
                          y = temperature_C,
                          color = lakename)) + 
  geom_point(size = 0.5, alpha = 0.5) + 
  starwars + # Add my theme
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              alpha = 0.5) +
  ylim(0, 35) + # Limit temperature to 35 degrees C
  xlab("Lake Depth (m)") +
  ylab("Temperature (degrees C)") + 
  ggtitle("July Temperatures by Depth, by Lake")

print(tempdepth.plot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake.anova1)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Paul Lake and Ward Lake have the same mean temperature as Peter Lake (e.g. the difference between their means is not statistically significant). None of the lakes are statistically distinct from each other lake.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
We could run a two-sample T-test on Peter Lake and Paul Lake. This would test against the null hypothesis that they have the same mean temperature.

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

Lake.pipe2 <- 
  Lake.pipe %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

temp.twosample <- t.test(Lake.pipe2$temperature_C ~ Lake.pipe2$lakename)
temp.twosample
  
```

>Answer: 
The two-sample t-test found no significant difference between the mean temperatures of Ward Lake and Crampton Lake. The p-value is 0.2649, well over the threshold for significance. We can therefore consider their mean temperatures equal. This is consistent with my results to question 16, which found no difference between these lakes. The adjusted p-value for their difference in the Tukey HSD test was 0.9714459, which is quite far from significant.
