---
title: "Assignment 3: Data Exploration"
author: "Annabelle White"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
getwd() # Check working directory
setwd("C:/Users/ardwh/OneDrive/Documents/NSOE/env872/EDA-Spring2023")
library(tidyverse)
library(lubridate) # Load packages
Neonics <- read.csv("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = T)
Litter <- read.csv("Data/Raw/NIWO_Litter/NEON_NIWO_Litter_massdata_2018-08_raw.csv", 
                   stringsAsFactors = T)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Neonicotinoids are applied as seed treatments and may percolate through soil, groundwater, and nearby plants. If these chemicals prove toxic to various insect taxa, they could have widely spread deleterious impacts on decomposers, pollinators, and all other insects.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Litter and woody debris are crucial to cycling nutrients into forest soils via decomposition. They have a pronounced impact on carbon and nitrogen cycling.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Tower plot and litter trap distributions differ between forested (20 plots, random traps) and low-stature vegeration areas (30 plots, targeted traps).
 2. Plot centers must be >50m from paved roads, and plot edges must be >10m from dirt roads.
 3. Deciduous forests are sampled every 2 weeks, but discontinued for up to 6 months in the dormant season. Evergreen forests are sampled every 1-2 months.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dimensions of bug data}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Effect summary}
summary(Neonics$Effect)
```

> Answer:
The most common effects are population (1803 studies) and mortality (1493 studies), several times more than the next most common effect (behavior). These make sense to study, as they can be surveyed through relatively uncomplicated methods, and give an overall snapshot of insect wellbeing. Trends in population and mortality would clearly indicate whether neonicotinoids are harming insect communities.

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Species list}
sort(summary(Neonics$Species.Common.Name))
```

> Answer:
The most commonly studied are species of bees and wasps. This makes sense, as bees and wasps are pollinators and provide an ecosystem service that is valuable to human food supply chains - including the agricultural fields that are employing neonicotinoids. If neonicotinoids are found to harm pollinators, this would have severe ramifications for their use.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Concentration class}
class(Neonics$Conc.1..Author.)
```

> Answer:
These data are not numeric because they contain characters such as >, /, and ~, which R does not process as numeric. This is good, because the data are not intended to be a continuous variable; they are arranged in categorical levels, comparing the differing levels of concentrations that each study used.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Test.Location graph}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year), bins = 25) # Looks nicer this way
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Test.Location colored graph}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 25)
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:
The most common test locations are in natural field settings and lab settings. Which of these is most common varies, with spikes in certain periods. For example, in the 2010s, lab tests increased to drastically more than any other test location, comprising most of the tests. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Endpoint bar graph}
ggplot(Neonics) +
	geom_bar(aes(x = Endpoint)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # Tilt labels
```

> Answer:
The most common endpoints are Lowest Observed Effect Level (LOEL), in which the lowest dose of concentration produced effects significantly different from control; and No Observed Effect Level (NOEL), in which the highest dose of concentration did not produce such effects. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Collection dates}

# Fix data type
class(Litter$collectDate) # Not a date; displays as factor (originally)
summary(Litter$collectDate) # To show the date format; it's year-month-day
Litter$collectDate <- ymd(Litter$collectDate) # Perform Y/M/D function on the vector
class(Litter$collectDate) # Check class; it's a date!
# Check for all unique dates:
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Number of plots}
unique(Litter$plotID)
```

> Answer:
summary() shows the distribution of values in a vector. unique() shows the number of distinct values in a vector. You could get the same information from summary() if you manually counted the display, but it seems foolish to do that in a coding course, and utterly unfeasible for larger datasets.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Group bar graph}
ggplot(Litter) +
	geom_bar(aes(x = functionalGroup)) + 
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5, hjust=1)) # Tilt labels
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Mass by Group}
ggplot(Litter) +
	geom_boxplot(aes(x = functionalGroup, y = dryMass, group = cut_width(functionalGroup, 1))) + 
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5, hjust=1)) # we slidin

ggplot(Litter) +
	geom_violin(aes(x = functionalGroup, y = dryMass)) + 
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5, hjust=1))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:
Because there is a very even distribution of masses across most functional groups, there is little variation to be displayed on the violin plot. As a result, the violins have been gruesomely flattened into single strings. As any orchesta concertgoer can attest, it is much more difficult to interpret the outputs from a single string than from a healthy violin. The boxplot is more effective because it shows that there is very little variation in mass for most functional groups.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
The highest biomass comes from needle litter and, to a lesser extent, mixed litter. This makes intuitive sense, as anyone who's ever set foot in a pine stand can tell you that litter is DENSE. There is a spot off a trail in the Duke Forest that I think would be a perfect place to take a nap in the dappled mid-morning sun. I crave it.
